Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics

### 2019-01-08 scagnostics.Rmd
Some rewording of the discussion.

Physics examples motivate us to investigate projection pursuit indexes that detect complex patterns in two dimensional projections. In this context we have investigated the use of indexes developed for variable selection as projection pursuit indexes. Having outlined a set of criteria that define a good index function, we have established methods for the evaluation of the index properties: smoothness, squintability, flexibility, rotation invariance and speed. To this end we used three different sets of simulated data, defined to have different underlying noise distributions (uniform or normal) and different structures (circle, sine and spiral), and we have evaluated several index functions proposed in the literature for variable selection (scagnostics indexes, maximum information based indexes, distance correlation and a spline based index). While all index functions are found to fall short on at least some of the criteria, we have shown that combined with the optimisation available via the guided tour they can already be used to identify interesting views, in particular the built-in structures in the simulated data sets.

We have also sketched some potential improvements of the indexes considered, in particular tuning parameters inherent to the index definition, such as binning, or different smoothing methods for the scagnostics indexes.

@@

We conclude that these methods are promissing and may yield new insight in the inter-dependence of model parameters in the light of experimental data ... other applications can also be envisioned??

There are three frontiers for future work building on our study. First, while we found the considered index functions useful to detect structures, several shortcomings have also been pointed out. While many definitions to identify clusters have been considered, there is much potential for the development of index functions that detect the types of structures considered here. Second, an implementation of interpolations from basis to basis, alleviating the need for rotation invariant indexes is envisioned. Finally, more generally, improved optimisation algorithms are required, e.g. for the identification of structures with small squint angle in high dimensions.

>>>>

Physics examples motivate us to investigate projection pursuit indexes that detect complex patterns in two dimensional projections. In this context we have investigated the use of indexes developed for variable selection as projection pursuit indexes. The set of relevant criteria that define a good index function includes: smoothness, squintability, flexibility, rotation invariance and speed. To evaluate these criteria we used three different sets of simulated data, defined to have different underlying noise distributions (uniform or normal) and different structures (circle, sine and spiral which can be considered as large, medium and fine structure respectively). The properties were assessed by a list of procedures introduced in Section (XXX cross reference?) and we have evaluated several index functions proposed in the literature for variable selection: scagnostics indexes, maximum information based indexes, distance correlation and a spline based index. While all index functions are found to fall short on at least some of the criteria, we have shown that combined with the optimisation available via the guided tour they can already be used to identify interesting views, in particular the built-in structures in the simulated data sets.

@@

The results are encouraging, showing large potential for discovering unanticipated relations between multiple variables. On the other hand there is much room for improvement on several fronts, as identified in this work.

Concretely we see three frontiers for future work building on our study. First, while we found the considered index functions useful to detect structures, several shortcomings have also been pointed out. While many definitions to identify clusters have been considered, there is much potential for the development of index functions that detect the types of structures considered here. Second, an implementation of interpolations from basis to basis, alleviating the need for rotation invariant indexes is envisioned. Finally, more generally, improved optimisation algorithms are required, e.g. for the identification of structures with small squint angle in high dimensions.

As a final remark we point out that small changes to the implementation of the guided tour algorithm would already be helpful, and we plan on implementing the following: allowing for a stopping condition based on index value, and making the step size for the selection of the most promising direction an argument when calling the geodesic optimiser.
